---
title: "Installing / Updating R & RStudio"
author: "Department of Psychology<br> The University of Edinburgh"
output:
  html_document:
    theme: readable
    toc: true
    toc_depth: 1
---


R is the name of both:

- the R programming language, i.e. the language you use to tell the PC what computations to perform; and
- the R interpreter, which actually executes the computations written in the R language.

RStudio is a user-friendly interface around R that makes the programming experience much nicer. It bundles together 

- a file editor panel, 
- a console panel with the R interpreter, 
- an environment panel showing all the objects you have created
- a files and plots panel to see your PC files and the plots you create.


The programming language R comes with built-in functionality, like computing the square root with the `sqrt()` function. Over the years, R users have created lots of additional functionalities that is available on demand. If you particular application requires extra functionality not available with the basic R installation, you can add it by *installing packages*. A package is an additional bit of functionality and it needs to be installed only once to bring it from the web to your own R installation. However, it is also important to keep packages updated once or twice a year.


This document explains how to 

a) install R and RStudio, if you never used them before; 
b) update R and RStudio, if you have old versions;
c) how to update your R packages.


---


# A. Installing R and RStudio

## A.1 Installing R {.tabset .tabset-fade .tabset-pills}

### Windows PC

#### Uninstall any previous installations

If you have a previous R installation, please uninstall the following software from your PC if you have it:

- R
- Rtools

#### Install R

Download the installer: https://cran.r-project.org/bin/windows/base/R-4.4.0-win.exe

Open the installer and follow the instructions.

#### Install RTools

Download RTools:

- If you have a Windows 64-bit PC (most likely) click here: https://cran.r-project.org/bin/windows/Rtools/rtools43/files/rtools43-5958-5975.exe

- If you have a Windows 32-bit PC (if your PC is quite old) you may need to find an older version here: https://cran.r-project.org/bin/windows/Rtools/

Open the installer and follow the instructions.



### Apple macOS PC

#### Uninstall any previous installations

If you have a previous R installation, please go to the Applications folder in your Finder window and move the following applications to the Bin:

- R
- XQuartz


#### Install R

Check which Apple PC you have. On the top-left of your screen click the Apple logo, click About This Mac, note what the Processor line says. If it has the word "Intel", you want to follow 1 below, and 2 otherwise.

1. Download this if you have an Apple M1-M3 macOS PC: https://cran.r-project.org/bin/macosx/big-sur-arm64/base/R-4.4.0-arm64.pkg

2. Download this if you have an older Intel-based macOS PC: https://cran.r-project.org/bin/macosx/big-sur-x86_64/base/R-4.4.0-x86_64.pkg

Open the installer and follow the instructions.


#### Install XQuartz

Download XQuartz: [https://xquartz.macosforge.org/](https://xquartz.macosforge.org/). 

Open the installer and follow the instructions.



### Chromebooks

You cannot install R on Chromebooks. You need to access a version of R installed on a remote machine or server.



## A.2 Installing RStudio {.tabset .tabset-fade .tabset-pills}

### Windows PC {#install-win}

Go to <https://posit.co/download/rstudio-desktop/> and, under section "2: Install RStudio", click the big blue "Download RStudio Desktop" button.

Open the installer and follow the instructions.

### Apple macOS PC {#install-mac}

Go to <https://posit.co/download/rstudio-desktop/> and, under section "2: Install RStudio", click the big blue "Download RStudio Desktop" button.

Open the installer and follow the instructions.


### Chromebooks {#install-chrome}

You cannot install RStudio on Chromebooks. You need to access a version of R installed on a remote machine or server.


## {-}


---


# B. Updating R and RStudio

If your R version or packages are old, you might get errors or find that the code required to run some examples is different from the one provided.

Recall that R is the underlying programming language, while RStudio is just a wrapper around R providing a nicer user interface (text editor, file explorer and graphical window). Updating RStudio (the interface) will not update the R programming language which runs your code.  

## B.1 Check which R version you have

Check that your R version is up-to-date by typing into the R console:

```
version
```

After pressing Enter, you should see  something like this:

```
platform       << this will be different for each of you >>     
arch           << this will be different for each of you >>                  
os             << this will be different for each of you >>     
system         << this will be different for each of you >>        
status                                     
major          4                           
minor          4.0                         
<< and more lines which you can ignore... >>
```

Please check here: https://www.r-project.org/ to see what the most recent version is. If you have a different number for "major" than the most recent, it may well be time for an update.  


## 2. Update R {.tabset .tabset-fade .tabset-pills}

### Windows PC

Open RStudio, and type the following in the R console. Press Enter to run the code.

```{r eval=FALSE}
install.packages("installr")
installr::updateR()
```


### Apple macOS PC

To update R, you need to remove any old R installations and then reinstall R. 

First, go to Finder, click Applications, and drag and drop R to the bin.
Then, follow the instructions from [A.1](#install-mac) to reinstall it.


### Chromebooks

You cannot install R on Chromebooks. You need to access a version of R installed on a remote machine or server.

You still may want to update packages, though. Check [section C](#update-pkgs) for that.

## {-}


---


# C. Updating R packages {#update-pkgs}

Open RStudio. In the R console, type the following and press Enter:

```{r eval=FALSE}
options(pkgType = "binary")
update.packages(ask = FALSE)
```

<br>

