---
title: "Making Pretty Tables in RMarkdown"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'left')
```

```{r panelset, echo=FALSE}
xaringanExtra::use_panelset()
```


# Why?  

Default tables are not that pretty when compiled: 

```{r}
tipjoke <- read.csv('https://uoepsy.github.io/data/TipJoke.csv')
tiptable <- table(tipjoke$Card, tipjoke$Tip)
tiptable
```

The output of `table()` (above) can sometimes look neat enough, but data in other formats is hard to make look nice (it will just get printed as the object it is, e.g. a `tibble` below)  

```{r}
library(tidyverse)
pass_scores <- read.csv("https://edin.ac/2wJgYwL") 
pass_scores %>% group_by(school) %>%
  summarise(
    meanPASS = mean(PASS),
    sdPASS = sd(PASS)
  )
```

# Functions for pretty tables

::: {.panelset}

::: {.panel}
#### kable()

__knitr__'s `kable()` is a fairly simple way to make tables and data look a little nicer. Simply pass something to `kable()` and will usually spit out something nicer. You can find a guide here: https://bookdown.org/yihui/rmarkdown-cookbook/kable.html  
We can also add captions to our table!  


```{r}
library(knitr)
kable(tiptable, caption = "Oh what a lovely table")
```
  
The `kable()` function also has various useful bits of functionality such as rounding all numeric values to the same number of digits, or setting a value for which all NAs will be displayed.  

```{r}
options(knitr.kable.NA = "")

pass_scores %>% group_by(school) %>%
  summarise(
    meanPASS = mean(PASS),
    sdPASS = sd(PASS)
  ) %>% kable(., digits = 1)
```

:::





::: {.panel}
#### gt()

The __gt__ package has a lot of useful functionality, and works well with grouped data. The best resource for this is at https://gt.rstudio.com. `gt()` also allows us to specify captions. 

```{r}
library(tidyverse)
library(gt)
tipjoke %>% 
  count(Card, Tip) %>%
  gt(caption = "Another lovely table!")
```
  
You can shape your data in different ways before passing it to `gt()`.  
 
```{r}
tipjoke %>% 
  group_by(Card) %>%
  summarise(
    TipYes = sum(Tip==1),
    TipNo = sum(Tip==0)
  ) %>%
  gt()
```

```{r}
pass_scores %>% group_by(school) %>%
  summarise(
    meanPASS = mean(PASS),
    sdPASS = sd(PASS)
  ) %>% gt()
```


:::

::: {.panel}
#### pander()

Pander is nice and simple, and works on the idea of converting R output into neat markdown syntax, which then get compiled into nice looking tables.  
You can read up on some of the functionality here: https://cran.r-project.org/web/packages/pander/vignettes/pander.html 

```{r}
library(pander)
pander(tiptable)
```

```{r}
pass_scores %>% group_by(school) %>%
  summarise(
    meanPASS = mean(PASS),
    sdPASS = sd(PASS)
  ) %>% pander()
```


:::



:::






