---
title: "Introduction to R and RStudio"
author: "Edinburgh Stats Team"
date: "DPUK Spring Academy 2025"
output: 
    bookdown::html_document2:
        theme: readable
        self_contained: true
        toc: true
        toc_depth: 2
toc-title: "Topics"
---

```{r, include = FALSE}
knitr::opts_chunk$set(message = FALSE)
```

---

## Installing and Updating R, RStudio, and R packages

Detailed installation instructions can be found [on this webpage](https://uoepsy.github.io/scs/dpuk/install-update-r.html).

### Installing R

- **Windows:**
  - Uninstall any previous R and Rtools installations.
  - Download installer: https://cran.r-project.org/bin/windows/base/R-4.4.3-win.exe.
  - For RTools (64-bit): https://cran.r-project.org/bin/windows/Rtools/rtools44/files/rtools44-aarch64-6459-6401.exe; for older PCs, see available 32-bit versions.
- **Apple macOS:**
  - Remove any previous R and XQuartz installations from Applications.
  - Download installer for your processor:
    - If Apple M1-M4: https://cran.r-project.org/bin/macosx/big-sur-arm64/base/R-4.4.3-arm64.pkg.
    - If Intel-based: https://cran.r-project.org/bin/macosx/big-sur-x86_64/base/R-4.4.3-x86_64.pkg.
  - Install XQuartz: https://xquartz.macosforge.org/.
- **Chromebooks:**
  - R cannot be installed; use Posit Cloud (https://posit.cloud/).

### Installing RStudio

- **Windows & macOS:**
  - Download RStudio Desktop from https://posit.co/download/rstudio-desktop/ and follow the installation instructions.
- **Chromebooks:**
  - RStudio is unavailable; use Posit Cloud (https://posit.cloud/).

### Updating R
- **Windows:**
  - In RStudio run:
    ```r
    install.packages("installr")
    installr::updateR()
    ```
- **Apple macOS:**
  - Remove the old R from Applications and reinstall using the appropriate installer.
  
### Updating RStudio
- Check via Help > Check for updates. If an update is available, uninstall RStudio, and then install RStudio again.

### Updating R Packages
- In RStudio run:

    ```r
    options(pkgType = "binary")
    update.packages(ask = FALSE)
    ```

---

## Introduction to R and RStudio

RStudio Panes and Interface

- **Console:** Execute commands interactively. To execute code, press Enter.
- **Editor:** Write, modify, and save your R code. Press `Ctrl+Enter` or `Cmd+Enter` to send code to the Console for execution.
- **Environment:** View your current objects and variables.
- **Files, Plots, Packages, and Help:** Access your files, see your plots, and access the help page. Install/update packages (more on this later).

Console vs R scripts vs Rmarkdown (.Rmd) files

- **Console:** The Console is used for running commands interactively. It is great for quick calculations or testing small pieces of code, but it does not save your work. Once you close RStudio, the commands in the Console are lost unless explicitly saved elsewhere.
- **R Scripts (.R files):** R scripts are plain text files where you can write and save R code. They are ideal for creating reusable code and documenting your workflow. You can execute code from an R script by sending it to the Console (e.g., using `Ctrl+Enter` or `Cmd+Enter`).
- **RMarkdown (.Rmd files):** RMarkdown files combine code, text, and output in a single document. They are used for creating dynamic reports, presentations, or documents that include both analysis and narrative. You can execute code chunks within an RMarkdown file and render the document into formats like HTML, PDF, or Word.

Code vs Comments (vs Text in Rmarkdown)

- **Code:** Code is the actual R commands that are executed to perform tasks. In R scripts, code is written directly in the file. In RMarkdown, code is written inside code chunks (e.g., ` ```{r} ... ``` `).

- **Text in RMarkdown:** In RMarkdown, text outside of code chunks is treated as narrative or explanatory text. It is written in Markdown syntax and is used to provide context, explanations, or documentation alongside the code and its output.

- **Comments:** Comments are lines of text in R scripts (or in RMarkdown code chunks) that are not executed as code. They are used to explain or document the code. In R, comments start with `#`. For example:
  ```r
  # This is a comment
  x <- 5  # Assign 5 to x
  ```

---

## Basic R Functionality

### Arithmetic and Calculations

R can perform basic arithmetic operations. Here are some examples:

```{r}
1 + 2   # Addition
5 - 3   # Subtraction
2 * 3   # Multiplication
1 / 2   # Division

# Exponentiation
2^3
# Square roots
sqrt(4)
9^(1/2)

# Standard functions such as log(), exp(), log10() also exist
```

Remember, order of operations matters! Use parentheses to ensure the correct order.

```{r}
# This will give different results
(1 + 2) * 3   # Parentheses first
1 + 2 * 3     # Multiplication first
```

---

### Getting help in R

```r
?sqrt
help(sqrt)
```
---

### Logical values

```{r}
# Logical
TRUE
FALSE
!TRUE             # not operator
2 == 3            # equal to operator
2 != 2            # not equal to operator
!(2 == 2)
2 > 3
2 <= 3
(2 > 1) & (2 < 3) # and operator
(2 > 1) | (2 < 3) # or operator
```

---

### Variables and Assignment

Variables are used to store values. 

Use `<-` or `=` for assignment into a name. 

You can later retrieve the value by calling the variable name.

```{r}
x <- 5          # Assign 5 to x
y <- 10         # Assign 10 to y
total <- x + y  # Add x and y
total           # Output 15
```

---

## Vectors

Vectors are one-dimensional arrays that can hold numeric, character, or logical data. You can create vectors using the combine function, `c()`.

```{r}
# Creating a numeric vector
numbers <- c(1, 2, 3, 4, 5)

# Creating a character vector
characters <- c("apple", "banana", "cherry")

# Creating a logical vector
logicals <- c(TRUE, FALSE, TRUE)

# Accessing elements of a vector
numbers[1]       # First element
characters[2:3]  # Second and third elements
logicals[c(1, 3)] # First and third elements

# Vectorised operations
numbers * 2       # Multiply each element by 2
numbers + 10      # Add 10 to each element
numbers > 3       # Logical comparison

# Member of a vector
"cherry" %in% characters
```

---

## Installing packages

In console:

````r
install.packages("palmerpenguins")
install.packages("tidyverse")
````

Using the Packages tab in RStudio.

---

## Introduction to tidyverse

This is a brief overview of tidyverse, the set of packages for data science in R.

```{r}
# Load the tidyverse collection of packages
library(tidyverse)
```

`tidyverse` is a collection of packages which includes:

- `ggplot2` for data visualisation
- `dplyr` for data manipulation
- `tidyr` for data tidying
- `readr` for data import
- `tibble` for data frames
- `stringr` for string manipulation
- `forcats` for categorical variables

and more... 

The good news is: you don't need to remember which package a function comes from. 
When you load the tidyverse, all of these packages are loaded at once.

You can find helpful cheatsheets [here](https://posit.co/resources/cheatsheets/).

---

## palmerpenguins: A tidyverse data example

A demonstration using the `penguins` dataset, provided by the [palmerpenguins package](https://allisonhorst.github.io/palmerpenguins/index.html).

The `penguins` dataset contains 344 rows and 8 variables:

- `species` - a factor denoting penguin species (Adélie, Chinstrap and Gentoo)
- `island` - a factor denoting island in Palmer Archipelago, Antarctica (Biscoe, Dream or Torgersen)
- `bill_length_mm` - a number denoting bill length (millimeters)
- `bill_depth_mm` - a number denoting bill depth (millimeters)
- `flipper_length_mm` - an integer denoting flipper length (millimeters)
- `body_mass_g` - an integer denoting body mass (grams)
- `sex` - a factor denoting penguin sex (female, male)
- `year` - an integer denoting the study year (2007, 2008, or 2009)

```{r}
library(palmerpenguins)
library(tidyverse)

# Look at the first six rows of the dataset
head(penguins)

# Structure of dataset
str(penguins)

# Summary of dataset
summary(penguins)

# A scatterplot exploring penguin body mass vs flipper length
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(color = "darkblue", alpha = 0.6) +
  labs(title = "Penguins: Flipper Length vs Body Mass",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)") +
  geom_smooth(method = "lm", color = "red", se = FALSE)

# A boxplot to explore the association between body mass and species
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  labs(title = "Penguins: Body Mass by Species",
       x = "Species",
       y = "Body Mass (g)")
```


---

## Descriptive statistics

## Computing Descriptive Statistics for a Variable

To compute the number of observations (`n`), mean (`M`), and standard deviation (`SD`) for a variable, you can use the following code:

```{r}
# Descriptive statistics for body_mass_g
penguins |>
  summarise(
    n = n(),
    M = mean(body_mass_g, na.rm = TRUE),
    SD = sd(body_mass_g, na.rm = TRUE)
  )
```

To get the same statistics for a variable grouped by another variable (e.g., species), you can use the `group_by()` function:

```{r}
# Descriptive statistics for body mass by species
penguins |>
  group_by(species) |>
  summarise(
    n = n(),
    M = mean(body_mass_g, na.rm = TRUE),
    SD = sd(body_mass_g, na.rm = TRUE)
  )
```

---

## Data subsetting

````r
penguins[1, ]           # First row

penguins[, 1]           # First column - by number
penguins[ , "species"]  # First column - by name
penguins$species        # First column - by name

penguins[c(1,2,3), ]    # First three rows

penguins[, c(1,2,3)]    # First three columns - by number
penguins[, c("species", "island", "bill_length_mm")] # First three columns - by name

penguins[c(1,2,3), c(1,2,3)] # First three rows and columns and range shortcut
penguins[1:3, 1:3] # Shortcut : for range of consecutive values
````

Pipe operator: take what's on the left and use it as the first argument of the function on the right.

```
take_this |>
  and_then_do_this()
```

````r
penguins |>
  slice(1:3)
````

````r
# Subset the data for only Adelie species
adelie_penguins <- penguins |> 
  filter(species == "Adelie")

# Remove specific rows of the penguins dataset
!is.na(penguins$body_mass_g)    # returns TRUE/FALSE for each row

# Filter out rows with NA values in body_mass_g
penguins_no_na <- penguins |> 
  filter( !is.na(body_mass_g) )

# Only keep columns species, body_mass_g
penguins_subset <- penguins_no_na |> 
  select(species, body_mass_g)
````


---

## Creating new columns

You can create new columns in a data frame using the `mutate()` function from the `dplyr` package from `tidyverse`.

The syntax works as follows:

```
data_name |>
  mutate(new_column_name = expression)
```

For example:

```{r}
# Create a new column for body mass in kg
penguins <- penguins |>
  mutate(body_mass_kg = body_mass_g / 1000)

# Create a new column for bill length and depth in cm
penguins <- penguins |>
  mutate(
    bill_length_cm = bill_length_mm / 10,
    bill_depth_cm = bill_depth_mm / 10
  )
```

If you want to overwrite an existing column, you can do so by using the same name in the `mutate()` function.


---

## Factors

Sometimes categorical variables are not stored as characters/factors. They could be stored as numbers which represent categories.

For example, `species` is a categorical variable with three levels: Adélie, Chinstrap, and Gentoo. Suppose we have a dataset where `species` is stored as numbers (1, 2, 3) instead of characters. This can lead to confusion when analyzing the data, as the numerical representation does not convey the actual categories. Furthermore, you can't take the mean of those values.

To convert a numeric variable to a factor, you can use the `factor()` function.


```{r}
# Toy dataset with species as number
data_example <- tibble(species = c(1, 2, 3, 1, 2, 3))
data_example

mean(data_example$species) # Non-sensical mean. You can't take the mean of a categorical variable.

# Convert species to a factor
data_example <- data_example |>
  mutate(species = factor(species))

mean(data_example$species) # R tells you you can't do it

# Convert species to a factor and choose order of levels
data_example <- data_example |>
  mutate(
    species = factor(species,
                     levels = c(3, 2, 1))
  )

# Convert species to a factor and use better labels - check the data codebook for the labels
data_example <- data_example |>
  mutate(
    species = factor(species,
                     levels = c(3, 2, 1),
                     labels = c("Gentoo", "Chinstrap", "Adelie"))
  )

# What are the factor levels?
levels(data_example$species)
```

---

## Reading data into R

Consider [this dataset](https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv)

A file path is a string that specifies the location of a file or directory in your computer's file system. It tells your system how to navigate through folders to locate the file. There are two main types:

**Absolute Path:** Specifies the complete directory list from the root folder.

**Relative Path:** Specifies the location relative to the current file or folder.

Paths are essential for accessing or referencing files correctly in your code.

````r
# Read from local file
penguins_data <- read_csv("path/to/penguins.csv")

# Read from URL
penguins_data <- read_csv("https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv")
````

The following functions are also available to read other types of files:

- `read_csv()` for CSV files
- `read_tsv()` for tab-separated files
- `read_excel()` for Excel files
- `library(haven)` and then
  - `read_dta()` for Stata files
  - `read_sav()` for SPSS files
  - `read_sas()` for SAS files

---

## Summary & Questions

This introductory session builds the foundation in R and tidyverse for our forthcoming four days which will focus on the linear model. 

- R & RStudio installation and interface.
- Basics of R with emphasis on tidyverse.
- A practical data example with palmerpenguins.
- Any questions?
