---
title: "<b> Introduction to the Linear Model </b>"
subtitle: "DPUK Spring Academy<br><br> "
author: "Josiah King, Umberto Noe, (and credits to Tom Booth)"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "April 2025"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    self_contained: true
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.height = 6)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #  base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
    base_color = "#95A5A6", #intro
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)

library(tidyverse)
library(kableExtra)
library(car)

theme_set(theme_gray(base_size = 15))
```


# Overview

- Day 1: What is a linear model?
- Day 2: But I have more variables, what now?
- Day 3: Interactions
- Day 4: Is my model any good?



---
class: center, middle
# Day 4
**Is my model any good?**

---
class: inverse, center, middle

<h2>Part 1: Key model assumptions </h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Basic model diagnostics </h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: What we have not covered? </h2>

---
# Linear model assumptions 
+ So far, we have discussed evaluating linear models with respect to:
	+ Overall model fit ( $F$ -ratio, $R^2$)
	+ Individual predictors

+ However, the linear model is also built on a set of assumptions.

+ If these assumptions are violated, the model will not be very accurate.

+ Thus, we also need to assess the extent to which these assumptions are met.


---
# Some data for today

.pull-left[
+ Let's look again at our data predicting salary from years or service and performance ratings (no interaction).

$$y_i = \beta_0 + \beta_1 x_{1} + \beta_2 x_{2} + \epsilon_i$$

+ $y$ = Salary (unit = thousands of pounds ).

+ $x_1$ = Years of service.

+ $x_2$ = Average performance ratings.
 
]

.pull-right[

```{r, echo=FALSE}
salary2 %>%
  slice(1:10) %>%
  kable(.) %>%
  kable_styling(full_width = F)
```


]

---
# Our model
```{r}
m1 <- lm(salary ~ perf + serv, data = salary2)
```

+ We will run all our assumptions based on the object `m1`

---
# Visualizations vs tests
+ There exist a variety of ways to assess assumptions, which broadly split into statistical tests and visualizations.

+ We will focus on visualization:
	+ Easier to see the nature and magnitude of the assumption violation
	+ There is also a very useful function for producing them all.

+ Statistical tests often suggest assumptions are violated when problem is small.
  + This is to do with the statistical power of the tests.
  + Give no information on what the actual problem is.
  + A summary table of tests will be given at the end of the lecture.


---
# Visualizations made easy
+ For a majority of assumption and diagnostic plots, we will make use of the `plot()` function.
  + If we give `plot()` a linear model object (e.g. `m1` or `m2`), we can automatically generate assumption plots.

+ We will also make use of some individual functions for specific visualizations.

+ Alternatively, we can also use `check_model()` from the `performance` package.
  + This provides `ggplot` figures as well as some notes to aid interpretation.
  + Caution that these plots are **not in a format to use directly in reports**

---
#  Linearity 
+ **Assumption**: The relationship between $y$ and $x$ is linear.
  + Assuming a linear relation when the true relation is non-linear can result in under-estimating that relation


+ **Investigated with**:
  + Scatterplots with loess lines (single variables)
  + Component-residual plots (when we have multiple predictors)


---
# Linear vs non-linear

.pull-left[

```{r, echo=FALSE, message=FALSE}
df2 <- tibble(
  x = rnorm(1000, 10, 2),
  y = 5 + .8*x + rnorm(1000, 0,.5),
  y2 = 5 + .6*(x^3) + rnorm(1000, 0,10)
)

df2 %>%
  ggplot(., aes(x=x, y=y)) +
  geom_point() +
  geom_smooth(method = "lm", se=F) +
  labs(x= "X", y="Y", title = "Linear Relationship")

```

]

.pull-right[

```{r, message=FALSE, echo=FALSE}
df2 %>%
  ggplot(., aes(x=x, y=y2)) +
  geom_point() +
  geom_smooth(method = "lm", se=F) +
  labs(x= "X", y="Y", title = "Non-linear Relationship")
```

]

---
#  What is a loess line?

+ Method for helping visualize the shape of relationships:

+ Stands for...
  + **LO**cally
  + **E**stimated
  + **S**catterplot
  + **S**moothing

+ Essentially produces a line with follows the data.

+ Useful for single predictors.

---
# Visualization

.pull-left[
```{r, warning=FALSE}
lin_m1 <- salary2 %>%
  ggplot(., aes(x=serv, y=perf)) +
  geom_point()+
  geom_smooth(method = "lm", se=F) + # <<
  geom_smooth(method = "loess", se=F, #<<
              col = "red") +
  labs(x= "Years of Service", y="Performance", 
       title = "Scatterplot with linear (blue) 
       and loess (red) lines")
```
]

.pull-right[
```{r, echo=FALSE, warning=FALSE, message=FALSE}
lin_m1
```

]

---
#  Non-linearity

+ With multiple predictors, we need to know whether the relations are linear between each predictor and outcome, controlling for the other predictors

+ This can be done using **component-residual plots**
  + Also known as partial-residual plots
		
+ Component-residual plots have the $x$ values on the X-axis and partial residuals on the Y-axis

+ *Partial residuals* for each X variable are:

$$\epsilon_i + B_jX_{ij}$$

+ Where :
	+ $\epsilon_i$ is the residual from the linear model including all the predictors
	+ $B_jX_{ij}$ is the partial (linear) relation between $x_j$ and $y$

---
#  `crPlots()` 

+ Component-residual plots can be obtained using the `crPlots()` function from `car` package

```{r, eval = F, warning=FALSE}
m1 <- lm(salary ~ perf + serv, data = salary2)
crPlots(m1)
```

+ The plots for continuous predictors show a linear (dashed) and loess (solid) line

+ The loess line should follow the linear line closely, with deviations suggesting non-linearity

---
#  `crPlots()`
```{r, echo= F, warning=FALSE, fig.height=3.8}
m1 <- lm(salary ~ perf + serv, data = salary2)
crPlots(m1)
```


???
+ Here the relations look pretty good.

+ Deviations of the line are minor

---
# Normally distributed errors 
+ **Assumption**: The errors ( $\epsilon_i$ ) are normally distributed around each predicted value.

+ **Investigated with**:
  + QQ-plots
  +	Histograms

	
---
# Visualizations 
+ **Histograms**: Plot the frequency distribution of the residuals.

```{r, eval=FALSE}
hist(m1$residuals)
```

--

+ **Q-Q Plots**: Quantile comparison plots.
	+ Plot the standardized residuals from the model against their theoretically expected values.
	+ If the residuals are normally distributed, the points should fall neatly on the diagonal of the plot.
	+ Non-normally distributed residuals cause deviations of points from the diagonal.
		+ The specific shape of these deviations are characteristic of the distribution of the residuals.

```{r, eval=FALSE}
plot(m1, which = 2) #<<
```


---
# Visualizations

.pull-left[

```{r, echo=FALSE}
hist(m1$residuals)
```

]


.pull-right[

```{r, echo=FALSE}
plot(m1, which = 2) #<<
```

]


---
#  Equal variance (Homoscedasticity) 

+ **Assumption**: The equal variances assumption is constant across values of the predictors $x_1$, ... $x_k$, and across values of the fitted values $\hat{y}$
	+ Heteroscedasticity refers to when this assumption is violated (non-constant variance)

+ **Investigated with**:
  + Plot residual values against the predicted values ( $\hat{y}$ ).

---
#  Residual-vs-predicted values plot 
+ In R, we can plot the residuals vs predicted values using `residualPlot()` function in the `car` package.
  + Categorical predictors should show a similar spread of residual values across their levels
  + The plots for continuous predictors should look like a random array of dots
	  + The solid line should follow the dashed line closely

```{r, eval=FALSE}
residualPlot(m1)
```

+ We can also get this plot using:

```{r, eval=FALSE}
plot(m1, which = 1)
```

---
#  Residual-vs-predicted values plot 

.pull-left[
```{r, echo=FALSE}
residualPlot(m1)
```

]

.pull-right[
```{r, echo=FALSE}
plot(m1, which = 1)
```
]



---
#  Independence of errors 
+ **Assumption**: The errors are not correlated with one another

+ Difficult to test unless we know the potential source of correlation between cases.

+ Essentially, if a design is between person, we will assume the errors to be independent.


---
#  Multi-collinearity 
+ This is **not an assumption of linear model**, but it is something we need to consider. 
  + It sits between assumptions and case diagnostics.
  
+ Multi-collinearity refers to the correlation between predictors
  + We saw this in the formula for the standard error of model slopes for an `lm` with multiple predictors.

+ When there are large correlations between predictors, the standard errors are increased
	+ Therefore, we don't want our predictors to be too correlated

---
#  Variance Inflation Factor 
+ The **Variance Inflation Factor** or VIF quantifies the extent to which standard errors are increased by predictor inter-correlations

+ It can be obtained in R using the `vif()` function:

```{r}
vif(m1)
```

+ The function gives a VIF value for each predictor

+ Ideally, we want values to be close to 1

+ VIFs> 10 indicate a problem

---
#  What to do about multi-collinearity 

+ In practice, multi-collinearity is not often a major problem

+ When issues arise, consider:
	+ Combining highly correlated predictors into a single composite
		  + E.g. create a sum or average of the two predictors
	+ Dropping an IV that is obviously statistically and conceptually redundant with another from the model


---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Key model assumptions </h2>
<h2>Part 2: Basic model diagnostics  </h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: What we have not covered? </h2>

---
#  Three important features

+ Model outliers
	+ Cases for which there is a large discrepancy between their predicted value ( $\hat{y_i}$ ) and their observed value ( $y_i$ )

--

+ High leverage cases
	+ Cases with an unusual value of the predictor ( $x_i$ )

--

+ High influence cases
	+ Cases who are having a large impact on the estimation of model

---
#  Influence

+ High leverage cases, when they are also linear model outliers, will have high **influence**

+ Cases with high influence, have a strong effect on the coefficients

+ If we deleted such a case, the linear model coefficients would change substantially


---
# Influence

+ If a handful of influential cases are responsible for the linear model results, the conclusions might not generalise very well

+ Multiple ways to consider influence.
  + Here we will discuss Cook's distance.
  
+ Cook's Distance of a data point $i$ (can be written many ways):


$$D_i = \frac{(\text{StandardizedResidual}_i)^2}{k+1} \times \frac{h_i}{1-h_i}$$

---
#  Cooks Distance 
$$\frac{(\text{StandardizedResidual}_i)^2}{k+1} = \text{Outlyingness}$$


$$\frac{h_i}{1-h_i} = \text{Leverage}$$

+ So $D_i = \text{Outlyingness} \times \text{Leverage}$


+ Cook's distance refers to **the average distance the $\hat{y}$ values will move if a given case is removed.**
  + If removing one case changes the predicted values a lot (moves the regression line), then that case is influencing our results.

---
#  Cooks Distance 

+ Many different suggestions for cut-off's:
  + $D_i > 1$ 
  + $D_i > \frac{4}{n-k-1}$
  + Or size relative all values in data set

---
#  Cook's distance in R

.pull-left[
```{r}
salary2 %>%
  mutate(
    cook = cooks.distance(m1) #<<
  ) %>%
  filter(., cook > 4/(100-3-1)) %>% #<<
  kable(.)  %>%
  kable_styling(., full_width = F)
```

]

.pull-right[
```{r, echo=F}
plot(m1, which = 4)
```

]


---
# Influence of coefficients
+ Cook's distance is a single value summarizing the total influence of a case

+ In the context of a lm with 2+ predictors, we may want to look in a little more detail.

+ **DFFit**: The difference between the predicted outcome value for a case with versus without a case included

+ **DFbeta**: The difference between the value for a coefficient with and without a case included

+ **DFbetas**: A standardised version of DFbeta
  + Obtained by dividing by an estimate of the standard error of the regression coefficient with the case removed

---
#  COVRATIO 
+ Influence on standard errors can be measured using the **COVRATIO** statistic
	+ COVRATIO value <1 show that precision is decreased (SE increased)  by a case
	+ COVRATIO value >1 show that precision is increased (SE decreased) by a case

+ Cases with COVRATIOS $> 1+[3(k +1)/n]$ or $< 1-[3( k +1)/ n ]$ can be considered to have a strong influence on the standard errors

---
#  COVRATIO in R 
+ COVRATIO values can be extracted using the `covratio()` function:
+ We can extract these measures using the `influence.measures()` function


---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Key model assumptions </h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Basic model diagnostics </h2>
<h2>Part 3: What we have not covered? </h2>

---
# In short, quite a lot!

+ Coding schemes for categorical data
+ Coding specific comparisons
+ Interactions with categorical variables and 2+ levels
+ Detailed probing of interactions
+ Statistical tests for assumptions
+ Assumption corrections for assumption violations
+ Bootstrapping
+ Extended model diagnostics
+ Non-continuous outcome variables
+ ....

---
# Material links

+ All our lecture materials can be found [here](https://uoepsy.github.io/)

---
class: center, middle
# Thanks all!